---
title: "Exploratory Analysis"
output: html_notebook
---

```{r}
#Required libraries

# Tidyverse for data science and exploration
require(dplyr)
require(tidyr)
require(readr)
require(tibble)
require(stringr)
require(purrr)
require(forcats)
require(rlang)

# enhances tidyverse
require(tidylog) # additional logging
require(magrittr) # additional data pipe syntax


# for reading data in multiple formats
require(readxl)
require(haven)

# visual analysis
require(ggplot2)
require(GGally) # extensions to ggplot
require(gt) # well formatted tables
# client-side interactive publishable graphics
require(plotly)
require(leaflet)
require(crosstalk)
require(htmlwidgets)
# server-side interactive graphics
require(shiny)
require(shinyjs)
# Canned Interactive EDA 
require(ExPanDaR)


```
## Exploring KU Book Processing Charges
```{r  }
# read KU data frame
KUbpc.df <- read_csv("Public Data/openapc-de/data/bpc.csv")
# read DOAB metadata
source('~/monograph/Public Data/DOAB/doabingest.R')
# DOABmeta.df <- doabFetch()
```
```{r  }


head(KUbpc.df)
head(summary(KUbpc.df))

ggplot(data = KUbpc.df, aes(KUbpc.df$institution)) + geom_bar() 

ggplot(data = KUbpc.df, aes(KUbpc.df$euro)) + geom_histogram()

```
## General Exploratory Data Analysis
```{r  }

ggplot(data = KUbpc.df) + geom_bar(mapping = aes(x = KUbpc.df$doab))

# Date to Doab
date_doab <- KUbpc.df %>% ggplot(data = KUbpc.df, mapping = aes(x = KUbpc.df$period, colour = KUbpc.df$doab)) + geom_freqpoly(binwidth = 0.1)
ggplotly(date_doab)

# publisher_euro <- KUbpc.df %>% 
# ggplot(data = KUbpc.df, mapping = aes(x = KUbpc.df$publisher, colour = KUbpc.df$euro)) + geom_freqpoly(binwidth = 0.1)

# Institution to Euro
institution_euro <- KUbpc.df %>% ggplot(data = KUbpc.df, mapping = aes(x = KUbpc.df$euro)) + geom_freqpoly(mapping = aes(colour = KUbpc.df$institution), binwidth = 500)

ggplotly(institution_euro)

```
## Idea: Publishers vs. Charges
## Question: How do the top 25% of publishers divide up charges (in Euro)?
## Observation: Charges are grouped around ~2000 Euros and ~8000 Euros. 
```{r  }

publisher_counts <- KUbpc.df %>%
    group_by(publisher) %>%
    tally

sorted_counts = arrange(publisher_counts, desc(n))

total_n = sum(sorted_counts$n)
quarter_n = 0.25 * total_n
new_n = sum(sorted_counts$n[0:6])

sorted_counts %>% filter(n > 24)

# filtered <- filter(KUbpc.df$publisher %in% sorted_counts$publisher)

filtered <- filter(KUbpc.df, KUbpc.df$publisher == 'transcript Verlag' |
                     KUbpc.df$publisher == 'Duke University Press' |
                     KUbpc.df$publisher == 'University of Michigan Press' |
                     KUbpc.df$publisher == 'Manchester University Press' |
                     KUbpc.df$publisher == 'Pluto Press' |
                     KUbpc.df$publisher == 'Liverpool University Press')

head(filtered)

euro_publisher <- filtered %>% 
  ggplot(data = filtered, mapping = aes(x = filtered$publisher, y = filtered$euro), 
         aes(x = filtered$publisher, y = filtered$euro)) + 
  # geom_count(aes(color = ..n.., size = after_stat(prop), group = euro)) + 
  geom_count(aes(color = ..n.., group = euro)) + 
  scale_size_area(max_size = 10) + 
  theme(axis.text = element_text(size = rel(0.75))) +
  labs(title = "How Publishers Divide Charges", x = "Top 25% of Publishers", y = "Price (Euro)", color = 'Number of Copies') +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo", " "), width = 17))

# ggplot:
ggplotly(euro_publisher)

# crosstalk:
ft <- highlight_key(filtered)
gg_ft <- ggplot(data = ft, mapping = aes(x = filtered$publisher, y = filtered$euro)) + 
  geom_count(aes(color = ..n.., size = after_stat(prop), group = euro)) + 
  labs(title = "How Publishers Divide Charges", x = "Top 25% of Publishers", y = "Price (Euro)", color = 'Number of Copies') +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo", " "), width = 17))
cross_ft <- bscols(
  filter_select("publisher", "Select a publisher", ft, ~publisher),
  ggplotly(gg_ft, dynamicTicks = TRUE),
  widths = c(12, 12)
)

bscols(cross_ft)

# shared_euro_publisher <- SharedData$new(filtered)
# leaflet(shared_euro_publisher) %>% addMarkers()
# data.table::data.table(shared_euro_publisher)


```
## Idea: Publishers' Charges vs. Year/OA Type
## Sub-Question: What best explains the particular division of charges? (Year, OA Type)
## Observation: Charges seem to be grouped by type of OA business model. 
```{r  }

head(filtered)

# Does Type of OA impact the particular division of charges?

euro_oa_publisher <- filtered %>% 
  ggplot(data = filtered, mapping = aes(x = filtered$backlist_oa, y = filtered$euro), 
         aes(x = filtered$backlist_oa, y = filtered$euro)) + 
  geom_count(aes(color = ..n.., group = euro)) + 
  scale_size_area(max_size = 10) + 
  theme(axis.text = element_text(size = rel(0.75))) +
  labs(title = "How OA Impacts Price Division of Charges", x = "Type of OA", y = "Price (Euro)", color = 'Number of Copies')

# ggplot:
ggplotly(euro_oa_publisher)

# crosstalk:
ft <- highlight_key(filtered)
gg_ft <- ggplot(data = ft, mapping = aes(x = filtered$backlist_oa, y = filtered$euro)) + 
  geom_count(aes(color = ..n.., size = after_stat(prop), group = euro)) + 
  labs(title = "How OA Impacts Division of Charges", x = "Type of OA", y = "Price (Euro)", color = 'Number of Copies')
cross_ft <- bscols(
  filter_select("publisher", "Select a publisher", ft, ~publisher),
  ggplotly(gg_ft, dynamicTicks = TRUE),
  widths = c(12, 12)
)

bscols(cross_ft)


# Does Year impact the particular division of charges?

euro_year_publisher <- filtered %>% 
  ggplot(data = filtered, mapping = aes(x = filtered$period, y = filtered$euro), 
         aes(x = filtered$period, y = filtered$euro)) + 
  geom_count(aes(color = ..n.., group = euro)) + 
  scale_size_area(max_size = 10) + 
  theme(axis.text = element_text(size = rel(0.75))) +
  labs(title = "How Year Impacts Price Division of Charges", x = "Year", y = "Price (Euro)", color = 'Number of Copies')

# ggplot:
ggplotly(euro_year_publisher)

# crosstalk:
ft <- highlight_key(filtered)
gg_ft <- ggplot(data = ft, mapping = aes(x = filtered$period, y = filtered$euro)) + 
  geom_count(aes(color = ..n.., size = after_stat(prop), group = euro)) + 
  labs(title = "How Year Impacts Division of Charges", x = "Year", y = "Price (Euro)", color = 'Number of Copies')
cross_ft <- bscols(
  filter_select("publisher", "Select a publisher", ft, ~publisher),
  ggplotly(gg_ft, dynamicTicks = TRUE),
  widths = c(12, 12)
)

bscols(cross_ft)


```
## Idea: Publishers vs. OA
## Question: What type of business model do the top 25% publishers use?
## Observation: Most have a higher proportion of True (moved to OA from traditional publishing) than False (already published OA).
```{r  }

oa_type <- filtered %>% 
  ggplot(data = filtered, mapping = aes(x = filtered$publisher, colour = filtered$backlist_oa), fill = filtered$backlist_oa) +
  geom_bar(position = "fill", width = 0.7, fill="#EAEAEA") +
  labs(title = "Business Model OA for Publishers", x = "Top 25% of Publishers", y = "Proportion", color = 'Types of OA') +
  theme(axis.text = element_text(size = rel(0.75))) +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo", " "), width = 17)) +
  scale_color_brewer(palette = "Set1")

ggplotly(oa_type)

# crosstalk:
ft <- highlight_key(filtered)
oa_ft <- ggplot(data = ft, mapping = aes(x = ft$publisher, colour = ft$backlist_oa), fill = ft$backlist_oa) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Business Model OA for Publishers", x = "Top 25% of Publishers", y = "Proportion of Backlist OA", color = 'Types of OA') +
  theme(axis.text = element_text(size = rel(0.75))) +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo", " "), width = 17))
# cross_oa_ft <- bscols(
#   filter_select("publisher", "Select a publisher", ft, ~publisher),
#   ggplotly(oa_ft, dynamicTicks = TRUE),
#   # widths = c(12, 12)
# )

# bscols(cross_oa_ft)


```
## Idea: Publishers' OA vs. Year
## Question: Did OA business models of the top 25% publishers change per year?
## Observation:
```{r  }

oa_time <- function(pub_name) {
  pub_ft <- filter(filtered, filtered$publisher == pub_name)
  
  pub_oa <- pub_ft %>% 
    ggplot(data = pub_ft, mapping = aes(x = pub_ft$period, colour = pub_ft$backlist_oa), fill = pub_ft$backlist_oa) +
    geom_bar(position = "fill", width = 0.7, fill="#EAEAEA") +
    labs(title = paste(pub_name, "'s OA Through the Years", sep = ""), 
         x = "Years", y = "Proportion of Backlist OA", color = 'Types of OA') +
    theme(axis.text = element_text(size = rel(0.75))) +
    scale_x_discrete(limits=c(2017, 2018, 2019)) +
    scale_color_brewer(palette = "Set1")

  ggplotly(pub_oa)
  
}

top25_list = c("transcript Verlag", "Duke University Press", "University of Michigan Press", "Manchester University Press", "Pluto Press", "Liverpool University Press")

oa_time("transcript Verlag")

oa_time("Duke University Press")

oa_time("University of Michigan Press")

oa_time("Manchester University Press")

oa_time("Pluto Press")

oa_time("Liverpool University Press")

```
## Idea: Revenue vs. OA
## Question: What total revenue are publishers receiving each year?
## Observation: 
```{r  }

# Finding total revenue for each publisher

revenue_finder <- function(pub_name) {
  pub_filtered <- filter(filtered, filtered$publisher == pub_name)
  rev = sum(pub_filtered$euro)
}

revenue_df <- data.frame("publisher" = top25_list)
revenue_list <- c()

for (i in top25_list) {
  revenue_list<-c(revenue_list,revenue_finder(i))
}

revenue_df$revenue <- c(revenue_list)
print(revenue_df)

# ggplot:
publisher_revenue <- revenue_df %>%
  ggplot(data = revenue_df, mapping = aes(x = revenue_df$publisher, y = revenue_df$revenue), fill = revenue_df$revenue) +
  geom_col() +
  labs(title = "Total Revenue for Publishers", x = "Top 25% of Publishers", y = "Revenue (Euro)", color = 'Types of OA') +
  theme(axis.text = element_text(size = rel(0.75))) +
  scale_x_discrete(labels = function(x) str_wrap(str_replace_all(x, "foo", " "), width = 17)) +
  scale_fill_brewer(palette = "Set1")

ggplotly(publisher_revenue)


```
## Idea: Revenue vs. OA
## Question: What revenue are publishers receiving per year?
## Observation: 
```{r  }

# Finding total revenue for each publisher

revlist_2017 <- c()
revlist_2018 <- c()
revlist_2019 <- c()

for (name in top25_list) {
  pub_name <- filter(filtered, filtered$publisher == name)
  rev_2017 = sum(pub_name[pub_name$period == 2017,]$euro)
  revlist_2017 <- c(revlist_2017, rev_2017)
  rev_2018 = sum(pub_name[pub_name$period == 2018,]$euro)
  revlist_2018 <- c(revlist_2018, rev_2018)
  rev_2019 = sum(pub_name[pub_name$period == 2019,]$euro)
  revlist_2019 <- c(revlist_2019, rev_2019)
}

# want x axis to be year
# want y axis to be total revenue
# want publisher by year

revenue_df <- data.frame("publisher" = top25_list)
revenue_df$'2017' <- c(revlist_2017)
revenue_df$'2018' <- c(revlist_2018)
revenue_df$'2019' <- c(revlist_2019)

print(revenue_df)

# ggplot:
pub_year_revenue <- revenue_df %>%
  ggplot(data = revenue_df, mapping = aes(x = c('2017', '2018', '2019'), y = c(revenue_df$'2017', revenue_df$'2018', revenue_df$'2019'), group = revenue_df$publisher)) +
  geom_line() + geom_point() +
  labs(title = "Total Revenue for Publishers", x = "Year", y = "Revenue (Euro)", color = 'Publishers') +
  theme(axis.text = element_text(size = rel(0.75)))

ggplotly(pub_year_revenue)


```
### Comparison of charges by year and backlist
```{r}
# create faceted plot object
charges.plot <- KUbpc.df %>% ggplot(aes(euro))+geom_histogram(bins=6)+facet_grid(rows=vars(period), cols = vars(backlist_oa))


## Present as Standard plot
 plot(charges.plot)

# this plot will render publicly https://htmlpreview.github.io/?https://github.com/MIT-Informatics/monograph/blob/master/00%20EDA%20Start.nb.html

```
### Interactive charges exploration
```{r}
 ggplotly(charges.plot)
# https://mit-informatics.github.io/monograph/demo.html

```
```
### Interactive Dataset Exploration 
```
```{r}
KUbpc.df %>% ExPanD(df=.       ,title="KU Book Processing Charges",export_nb_option = TRUE)
# ExPanD uses shiny() which works running R locally, but isn't going to work through github. Could publish through shinyapps.io (low usage only), or export  a non-interactive notebook it
# see: https://drmaltman.shinyapps.io/demo/
```

